A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-07 paper.tex
Tidy up "history" section a bit

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately confusion exists over terminology and there is
little agreement on shared standards for data interchange,
significantly hampering progress in the field.
This lack of clarity and agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous,
% is it ambiguous? How? Can we simply delete this?
fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes and an edge
defines genetic inheritance between two nodes, with annotations on
each edge specifying which regions of the genome have
been inherited via that route.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be captured by a single tree: instead,
the inheritance of each section of the genome may be described
by a (slightly) different local tree~\citep{hudson1983properties}. Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning genealogy-based approaches have
not been widely used in population and statistical genetics. Until recently, analysis of
population genetic data has therefore primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the full underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the sample.
Numerous extensions to the model have been proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}, incorporating various types of evolutionary processes.
In particular, \citet{hudson1983properties}
analysed the effects of recombination on coalescent trees, generating several fundamental analytical results,
as well as describing the elegant algorithm
still at the heart of widely-used simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Crucially, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry
was viewed as a set of disconnected local trees rather than
as a single cohesive object.

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
formulating it as a stochastic process where each realisation
is a genealogy encoded as a graph,
and showed that the local trees generated by Hudson's algorithm
are embedded within this graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to both the stochastic process and
its random realisations as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to reason about and infer recombinant ancestry as a single object,
which was not possible within Hudson's framework.

>>>>

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
% https://www.overleaf.com/5594399339wckggzwhnfvv
\begin{abstract}
Recent breakthroughs have made it possible to infer genetic ancestry in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately confusion exists over terminology and there is
little agreement on shared standards for data interchange,
significantly hampering progress in the field.
This lack of clarity and agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous,
% is it ambiguous? How? Can we simply delete this?
fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes and an edge
defines genetic inheritance between two nodes, with annotations on
each edge specifying which regions of the genome have
been inherited via that route.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of a
phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be captured by a single tree: instead,
the inheritance of each section of the genome may be described
by a (slightly) different local tree~\citep{hudson1983properties}.
Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning genealogy-based approaches have
not been widely used in population and statistical genetics.
Until recently, analysis of
population genetic data has therefore primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the full underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the sample.
Numerous extensions to the model have been
proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
incorporating many evolutionary processes.
\citet{hudson1983properties}
first incorporated recombination into the coalescent process,
providing several fundamental analytical results
and describing the basic simulation algorithm, still in
widespread use~\citep{hudson2002generating,kelleher2016efficient,
baumdicker2021efficient}.
In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
formulating it as a stochastic process where each realisation
is encoded as a graph~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}.
They referred to both the stochastic process and
its random realisations as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
as originally described requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to reason about and infer recombinant ancestry as a single object,
which was not possible within Hudson's framework, which emphasised
the collection of local trees along the genome resulting from recombination.
